---
title: "Examples from dplyr with the athletes dataset"
author: "Steven Moran and Alena Witzlack-Makarevich"
date: "2024-07-23"
output: html_document
---

# Introduction

Let's use the data set `athletes.csv` for an example of working with some of `tidyverse` tools. 

First load the R libraries.

```{r message = FALSE}
library(tidyverse)
```

# Loading data and having a look

```{r message = FALSE}
athletes <- read_csv("data/athletes.csv")
```

Have a look.

```{r}
head(athletes)
```

```{r}
tail(athletes)
```

Have a look at the structure of the data with `str()`.

```{r}
str(athletes)
```

Character variables can converted into "factor" in R, i.e., categorical data. This makes working with them easier for various purposes, e.g., counting categories, doing certain statistics.

```{r}
athletes$gender <- as.factor(athletes$gender)
athletes$sport <- as.factor(athletes$sport)
athletes$country <- as.factor(athletes$country)
```

Now note the change in data type:

```{r}
str(athletes)
```

Another useful function is `summary()`:

```{r}
summary(athletes)
```


# Tidyverse functions for data wrangling

Now some functions so that we can select, filter, transform, extract, and summarize aspects of the data.

## `select()`

When there are two commands they are equivalent, i.e., you can either pass the dataframe as a parameter within the function (bolded):

* select(**athletes**, name, height, weight)

or you can "pipe" ("%>%"), i.e., give, the dataframe to the function:

* **athletes** %>% select(name, height, weight)

Piping allows you to string multiple functions together, e.g.:

* athletes %>% select(name, height, weight) %>% head()

```{r}
# These lines are equilvanent, i.e., they are two different ways of doing the same thing.

select(athletes, name, height, weight)

athletes %>% select(name, height, weight)
```

```{r}
# These lines are equilvanent, i.e., they are two different ways of doing the same thing.

select(athletes, age:weight)

athletes %>% select(age:weight)
```

```{r}
# These lines are equilvanent, i.e., they are two different ways of doing the same thing.

select(athletes, -birthdate, -age)

athletes %>% select(-birthdate, -age)
```

These functions do not modify the data.

```{r}
# These lines are equilvanent, i.e., they are two different ways of doing the same thing.

select(athletes, birthdate, age)

athletes %>% select(birthdate, age)
```

But you can save the results into a new data frame.

```{r}
# These lines are equilvanent, i.e., they are two different ways of doing the same thing.
athletes_age <- select(athletes, birthdate, age)
athletes_age

athletes_age <- athletes %>% select(birthdate, age)
athletes_age
```

## `arrange()`

`arrange()` changes the order of the rows.

```{r}
# These lines are equilvanent, i.e., they are two different ways of doing the same thing.
arrange(athletes, height, age)

athletes %>% arrange(height, age)
```

```{r}
# These lines are equilvanent, i.e., they are two different ways of doing the same thing.
arrange(athletes, desc(height), age)

athletes %>% arrange(desc(height), age)
```

```{r}
# These lines are equilvanent, i.e., they are two different ways of doing the same thing.
arrange(athletes, gender, desc(age))

athletes %>% arrange(gender, desc(age))
```


## `mutate()`

`mutate()` always adds new columns at the end of your data set.

First, create with `select()` a narrow data set `athletes_narrow`.

```{r}
# These lines are equilvanent, i.e., they are two different ways of doing the same thing.
athletes_narrow <- select(athletes, name, gender, age, sport, height, weight)
athletes_narrow

athletes_narrow <- athletes %>% select(name, gender, age, sport, height, weight)
athletes_narrow
```

Next, add the column BMI (body mass index). The BMI is calculated as the body mass (`weight`) divided by the square of the body `height`. It is universally expressed in units of kg/m^2^. 

```{r}
# These lines are equilvanent, i.e., they are two different ways of doing the same thing.
mutate(athletes_narrow, BMI = weight/height^2)

athletes_narrow %>% mutate(BMI = weight/height^2)
```

Notice that `mutate` does not overwrite the existing data frame.

```{r}
# These lines are equilvanent, i.e., they are two different ways of doing the same thing.

mutate(athletes_narrow, BMI = weight/height^2)
athletes_narrow

athletes_narrow %>% mutate(BMI = weight/height^2)
athletes_narrow
```

To add the new column to it permanently, you have to overwrite the original data frame:

```{r}
# These lines are equilvanent, i.e., they are two different ways of doing the same thing.

athletes_narrow <- mutate(athletes_narrow, BMI = weight/height^2)
athletes_narrow

athletes_narrow <- athletes_narrow %>% mutate(BMI = weight/height^2)
athletes_narrow
```


### `summarize()`

The dplyr function `summarize()` (or `summarise()`) summarizes multiple values in a single value. From now on we simply use the pipe "%>%".

```{r, eval = FALSE}
athletes %>% summarize(totals = sum(gold_medals))
```

## `group_by()`

Groups table data into groups on which some other operation may operate.

For example, an important function of `summarize()` is in coordination with the `group_by()` function. The dplyr `group_by` function take an existing data frame and performs an operation by group.

We can also use `group_by` on the athletes data. For example, how many gold medals per country?

```{r}
athletes %>% group_by(country) %>% summarize(gold_medals = sum(gold_medals))
```

Maybe for viewing purposes it's better to arrange them by number of gold medals instead of alphabetically by country name.

```{r}
athletes %>% group_by(country) %>% summarize(gold_medals = sum(gold_medals)) %>% arrange(desc(gold_medals))
```

```{r}
athletes %>% 
  group_by(country) %>% 
  summarize(gold_medals = sum(gold_medals)) %>% 
  arrange(desc(gold_medals))
```

## `filter()`

Probably the most useful `dplyr` function is `filter()`.
```{r}
filter(athletes_narrow, height < 2)
```

```{r}
filter(athletes_narrow, sport == "Basketball")
```

When you’re starting out with R, the easiest mistake to make is to use = instead of == when testing for equality. When this happens you’ll get an informative error. Try it out:

```{r, eval=FALSE}
filter(athletes_narrow, age = 15)
```


## Filtering NA

Another important function when filtering is to identify and potentially filter out `NA` cells. These are missing or unknown values in the data set. 

Let's filter the rows in `athletes` that do not have a height value, i.e., they are `NA` in the table.

```{r}
athletes %>% filter(is.na(height))
```

You can also filter to **remove** `NA`s, which is often useful for when you want to visualize the data. Use the logical operation `!` mentioned above, i.e., "not".

```{r}
athletes %>% filter(!is.na(height))
```

You can also combined the filters. For example, if you want all rows in `athletes` that do not have `NA` values for `height` and `weight`. Notice how the number of rows decreases.

```{r}
athletes %>% filter(!is.na(height)) %>% filter(!is.na(weight))
```

If you want to check if there are any `NA`s in a column, you can also use the `any()` function.

```{r}
any(is.na(athletes$height))
any(is.na(athletes$age))
```


# The table function

Another useful function is called `table()`. What does it do?

```{r}
table(athletes$sport)
```

Note you need to use the `exclude = FALSE` parameter, if you want the `table()` function to count `NA`s. How many athlete

```{r}
table(athletes$height)
table(athletes$height, exclude = FALSE)
```
