---
title: "Salos 2024 summer school: Exploring PHOIBLE"
author: "Steven Moran and Alena Witzlack"
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output:
  github_document:
      toc: true
---

------------------------------------------------------------------------

These are the R libraries you will need. Add more if you need them.

```{r, message=FALSE}
library(tidyverse)
```

# Introduction

The [PHOIBLE database](https://phoible.org) is an online repository of phonological inventory data. From the website one can browse data by various categories, e.g.:

-   [Inventories](https://phoible.org/inventories)
-   [Languages](https://phoible.org/languages)
-   [Segments](https://phoible.org/parameters)

# Load the data

First load the data. Use this version of the dataset available in this repository in the [phoible.csv](data/phoible.csv) in the data directory.

```{r}
# Insert your code below here

library(readr)
df <- read_csv("~/Library/CloudStorage/OneDrive-UniversityofHelsinki/internship_projects_summerschool/Salo Summer School/source_data/data/phoible.csv")
```

If you use the `read_csv` function from `tidyverse`, you will want to convert the character data (columns) into categorical ("factor") data in R. You can do this by uncommenting the code (delete the "\#") below and either renaming `df` to whatever you called your dataframe or loading the data with the variable `df` above.

```{r}
df$language <- as.factor(df$language)
df$genus <- as.factor(df$genus)
df$fam <- as.factor(df$fam)
```

How do we have a quick look at the dataset's structure? How many variables/columns there are? How many rows there are?

```{r}
# Insert your code below here
head(df)
```

How can we have a quick look at the first few rows (the "head") of the data?

```{r}
# Insert your code below here

```

How do we get a summary of the contents (i.e., each column) of the data frame?

```{r}
# Insert your code below here

```

# Exploring the data

We can use the `dplyr` functions to explore the data. See the [athletes file](athletes.md) for examples.

-   Select the language, genus, and phoneme count columns.

```{r}
# Insert your code below here

```

-   Arrange those columns by largest to smallest phoneme inventory size.

```{r}
# Insert your code below here

```

-   Which language has the most consonants?

```{r}
# Insert your code below here

```

-   Filter all languages that have more than 100 consonants.

```{r}
# Insert your code below here

```

-   Which language has the least number of vowels?

```{r}
# Insert your code below here

```

-   Filter all languages that have less than 3 vowels.

```{r}
# Insert your code below here

```

-   Which language has the most number of tones?

```{r}
# Insert your code below here

```

-   How many languages do not have tone?

```{r}
# Insert your code below here

```

-   How many languages are there in the sample?

```{r}
# Insert your code below here

```

-   How many language families are there in the sample?

```{r}
# Insert your code below here

```

-   Use the mutate function to determine the average (mean) number of phonemes across all languages in the sample.

```{r}
# Insert your code below here

```

-   Do the same for consonants, vowels, and tones.

```{r}
# Insert your code below here

```

-   Can you do the same "by group", i.e., by language family?

```{r}
# Insert your code below here

```

-   How about by language genus? (What's the difference?)

```{r}
# Insert your code below here

```

-   Are there any NAs in the table?

```{r}
# Insert your code below here

```

What other questions can you ask of the data with the tools that you've learned about?
