---
title: "Salos 2024 summer school: Working with PHOIBLE"
author: "Steven Moran and Alena Witzlack"
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output:
  github_document:
      toc: true
---

***

The libraries you will need.

```{r, message=FALSE}
library(tidyverse)
```


# Introduction

The [PHOIBLE database](https://phoible.org) is an online repository of phonological inventory data. Let's have a look at the website, where we can explore by various categories:

* [Inventories](https://phoible.org/inventories)
* [Languages](https://phoible.org/languages)
* [Segments](https://phoible.org/parameters)

***

But what if we want to explore and work with the data programmatically? What's the first thing we need to do?

***

We need to access the data. So, where is it?

One option is to download the data from the website:

* https://phoible.org/download

Data comes in different data formats (and different data types).

In linguistics, data are usually curated and we have data sets in different stages of development. For example, the PHOIBLE data are curated and maintained in a working format in a [GitHub](https://github.com) repository here:

* https://github.com/phoible/dev

This repository contains the raw data and code (scripts) to transform different data formats into a single spreadsheet:

* https://github.com/phoible/dev/blob/master/data/phoible.csv

This is a rather large spreadsheet, so we have made a smaller aggregated one here:

* https://github.com/bambooforest/salos/blob/main/data/phoible.csv

When a data table (aka spreadsheet) is smaller, GitHub will display it online for you:

* https://github.com/bambooforest/salos/blob/main/data/phoible.csv

***

How do we load the data into R/RStudio?

```{r}
df <- read_csv('data/phoible.csv')
df
```











