---
title: "Recreate PHOIBLE visualizations"
author: "Alena Witzlack-Makarevich and Steven Moran"
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output:
  html_document
---

```{r, message = FALSE, echo = FALSE}
phoible <- read.csv("data/phoible.csv")
```

```{r, message = FALSE, echo = FALSE}
library(tidyverse)
library(readxl)
require(forcats)
library(cowplot)
library(knitr)
library(kableExtra)
```

# Introduction

Here are a series of taks to recreate visualizations of phoible data (see [phoible.csv](data/phoible.csv)).


# Task 1: Compare Indo-European (ieur) vs. Nilo-Saharan (nsah)

Recreate the boxplot below that compares the number of phonemes in Nilo-Saharan (`nsah`) languages and in Indo-European languages (`ieur`). 

You need to filter for these two families first.

```{r, message = FALSE, echo = FALSE, out.width="60%"}
phoible %>% 
        filter(fam %in% c("nsah", "ieur")) %>%
        ggplot(aes(y = phonemes, x = fam, fill = fam)) +
        geom_boxplot()+
        labs(title="Number of phonemes in Indo-European and Nilo-Saharan languages",
             x ="family", y = "number of phonemes") +
        theme_light()+
        theme(axis.text.y = element_text(size = 12), axis.text.x = element_text(size = 12))
```

# Task 2: Add some jitter to Indo-European (ieur) vs. Nilo-Saharan (nsah)

Add a scatter plot atop of the boxplots to give some idea about the number of languages in each group.

```{r, message = FALSE, echo = FALSE, out.width="60%"}
phoible %>% 
        filter(fam %in% c("nsah", "ieur")) %>%
        ggplot(aes(y = phonemes, x = fam, fill = fam)) +
        geom_boxplot()+
        geom_jitter(alpha = 0.3)+
        labs(title="Number of phonemes in Indo-European and Nilo-Saharan languages",
             x ="family", y = "number of phonemes") +
        theme_light()+
        theme(axis.text.y = element_text(size = 12), axis.text.x = element_text(size = 12))
```

# Task 3: boxplot vs. density plot

Visualize the distribution of the number of phonemes in two language families with density plots. 

```{r, message = FALSE, echo = FALSE, out.width="60%"}
phoible %>% filter(fam %in% c("aust", "anes")) %>% 
  ggplot(aes(x = phonemes, fill = fam)) +
  geom_density(alpha = 0.3) + 
  labs(title="Number of phonemes in two language families") +
  theme_light()+
  theme(axis.text.y = element_text(size = 14), axis.text.x = element_text(size = 14))
```

# Task 4: histogram in facets

Visualize the distribution of phonemes in the three families with faceted histograms. Order them by the number of languages in each group from large to small.

```{r, message = FALSE, echo = FALSE, out.width="60%"}
phoible %>% filter(fam %in% c("ausa", "altc", "trng")) %>% 
  # change order of factor by the family size
  mutate(fam = fct_infreq(fam)) %>% 
  ggplot(aes(x = phonemes, fill = fam)) +
  geom_histogram(binwidth = 5) + 
  labs(title="Number of phonemes in three language families",
  x ="family", y = "number of phonemes") +
  theme_light()+
  theme(axis.text.y = element_text(size = 14), axis.text.x = element_text(size = 14)) +
  facet_grid(. ~ fam )
```

# Task 5: Scatterplot without much overlapping

Visualize the correlation of number of consonants and number of vowels.

```{r, message = FALSE, echo = FALSE, out.width="60%"}
phoible %>% filter(fam %in% c("ausa", "altc", "trng")) %>% 
  ggplot(aes(x = vow, y = con, col = fam)) +
  geom_jitter() + 
  labs(title="Number of consonants and vowels",
  x ="Number of vowels", y = "Number of consonants") +
  theme_light()+
  theme(axis.text.y = element_text(size = 14), axis.text.x = element_text(size = 14))
```

<!-- 
```{r, message = FALSE, echo = FALSE, out.width="60%"}
phoible %>% filter(fam %in% c("ausa", "altc", "trng")) %>% 
  ggplot(aes(x = vow, y = obs, col = fam)) +
  geom_jitter() + 
  geom_smooth() +
  labs(title="Number of consonants and vowels",
  x ="number of vowels", y = "number of consonants") +
  theme_light()+
  theme(axis.text.y = element_text(size = 14), axis.text.x = element_text(size = 14))
```
-->
