---
title: "Warriner 2013"
author: "Steven Moran and Alena Witzlack-Makarevich"
date: "2024-07-25"
output: github_document
---

```{r, message=FALSE}
library(tidyverse)
```

# Introduction

The idea here is to replicate the results of Warriner et al. (2013):

* https://doi.org/10.3758/s13428-012-0314-x

The abstract:

>>> Information about the affective meanings of words is used by researchers working on emotions and moods, word recognition and memory, and text-based sentiment analysis. Three components of emotions are traditionally distinguished: valence (the pleasantness of a stimulus), arousal (the intensity of emotion provoked by a stimulus), and dominance (the degree of control exerted by a stimulus).

First let's load the data.

```{r, warning=FALSE}
warriner <- read_csv("data/warriner2013.csv")
```

# Result 1: Valence and dominance 

Let's explore the relationship between valence and dominance visually.

```{r}
# the basic command looks like this: all variables you plot go into aes()
# these are you aesthetics, i.e. aspects of the data you want to see in the plot
ggplot(warriner, aes(y = Dominance_mean, x = Valence_mean)) +
  geom_point()
```

Then you can work on over plotting.

```{r}
# alpha stands for transparency (from 0 to 1)
ggplot(warriner, aes(y = Dominance_mean, x = Valence_mean)) +
  geom_point(alpha =0.2, size = 0.1, col = "dodgerblue3")
```

Now let's add a regression line.

```{r}
# by default, geom_smooth() adds a local regression line
ggplot(warriner, aes(y = Dominance_mean, x = Valence_mean)) +
  geom_point(alpha =0.2, size = 0.1, col = "dodgerblue3") +
  geom_smooth(col = "firebrick3")+
  theme_bw()
```

A linear regression needs `method = "lm"`.

```{r}
ggplot(warriner, aes(y = Dominance_mean, x = Valence_mean)) +
  geom_point(alpha =0.2, size = 0.1, col = "dodgerblue3") +
  geom_smooth(method = "lm", col = "firebrick3")+
  theme_bw()
```

It would be helpful to add a few words as examples.

```{r}
# looking up the coordinates for individual words
ggplot(warriner, aes(x = Valence_mean, y = Dominance_mean))+
  geom_point(alpha =0.1, size = 0.07, col = "dodgerblue3") +
  geom_smooth(method = "lm", col = "firebrick3")+
  annotate(geom="text", x=8.47, y=7.21, label="happy", angle = -70)+
  annotate(geom="text", x=6.05, y=3.29, label="lotto", angle = -70) +
  annotate(geom="text", x=7, y=5.73, label="dog", angle = -70)+
  theme_bw()
```

The relationship between valence and dominance is linear: words that make people feel happier also make them feel more in control (r = 0.717).

Calculate the slope and intercept, name your model e.g. m1.

```{r}
m_Dom_Val <- lm(warriner$Dominance_mean ~ warriner$Valence_mean)
```

Have a look at the intercept and slope.

```{r}
m_Dom_Val
```

Could you guess why the slope is different from what is reported in the paper? (Hint: it's a mistake in the paper!)

```{r}
# summary() provides more details, including e.g. Adjusted R-squared:  0.5135
summary(m_Dom_Val)
```

You can calculate the correlation coefficient with `cor()`.

```{r}
cor(warriner$Valence_mean, warriner$Dominance_mean)
```


# Result 2: Valence vs. arousal

Words that are very positive or very negative are more arousing than those that are neutral.

This is corroborated by the positive correlation between valence and arousal.

* for positive words (mean valence rating > 6; r = .273, p < .001) 
* and the negative correlation between valence and arousal 
* for negative words (mean valence rating < 4; r = −.293, p < .001)

# Task 2:

* use filter() to obtain two subsets of the data:
* one data set for low valence and one for high valence following the specifications above 

* produce two separate scatter plots
* address the overplotting issue

* calculate two separate models for the two subsets

* produce a plot with a linear regression, not too much overplotting and a few more words as examples

# Result 3: Valence vs. arousal

The relationship between arousal and dominance is also U-shaped as corroborated by the positive correlation between dominance and arousal for high-rated dominance words (mean rating > 6; r = .139, p < .001) and the negative correlation between dominance and arousal for low-rated dominance words (mean rating < 4; r = −.193, p < .001).

# Task 3:

* use filter() to obtain two subsets of the data:
* one data set for low valence and one for high valence following the specifications above 

* produce two separate scatter plots
* address the overplotting issue

* calculate two separate models for the two subsets

* produce a plot with a linear regression, not too much overplotting and a few more words
