---
title: "Salos 2024 summer school: What's the background of the class?"
author: "Steven Moran"
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output:
  github_document:
      toc: true
---

***

These are the R libraries you will need:

```{r, message=FALSE}
library(tidyverse)
library(knitr)
```

Load the data.

```{r, message=FALSE}
df <- read_csv('data/Background questionnaire (Answers) - Risposte del modulo 1.csv')
```

Have a look.

```{r}
df
```

Have a nicer look.

```{r}
kable(df)
```

What data types do we have? Each column represents a data type, e.g., numbers or categories.

```{r}
str(df)
```

Do you have programming experience?

```{r}
df$`Do you have any programming (coding) experience?`
```

How do we tally the counts?

```{r}
table(df$`Do you have any programming (coding) experience?`)
```

Let's visualize it.

```{r}
# Convert the responses to a data frame
response_df <- as.data.frame(table(df$`Do you have any programming (coding) experience?`))

# Bar plot
ggplot(response_df, aes(x=Var1, y=Freq, fill=Var1)) +
  geom_bar(stat="identity") +
  labs(title="Yes/No Responses", x="Response", y="Count") +
  scale_fill_manual(values=c("lightblue", "salmon"))
```

The dreaded pie chart.

```{r}
ggplot(response_df, aes(x="", y=Freq, fill=Var1)) +
  geom_bar(width=1, stat="identity") +
  coord_polar(theta="y") +
  labs(title="Yes/No Responses") +
  scale_fill_manual(values=c("lightblue", "salmon"))
```

If so, please rate your programming knowledge (1-5).

```{r}
df$`If so, please rate your programming knowledge`
```

We can look at the distribution. (But the numbers don't add up!)

```{r}
table(df$`If so, please rate your programming knowledge`)
```

How many people did not respond ("NA")?

```{r}
table(df$`If so, please rate your programming knowledge`, exclude=FALSE)
```

How about a quick summary of the numbers?

```{r}
summary(df$`If so, please rate your programming knowledge`)
```

Have you used R?

```{r}
table(df$`Have you used R?`)
```

Any R experts (1-5)?

```{r}
table(df$`If so, please rate your R expertise`, exclude=FALSE)
```

Let's see if anyone reported that they don't have coding experience, but they have used R. Heh?

```{r}
df %>% filter(`Do you have any programming (coding) experience?` == "No") %>%
  filter(`Have you used R?` == "Yes") %>%
  select(`Do you have any programming (coding) experience?`, 
              `Have you used R?`)
```



